An efficient composition of bidirectional programs by tupling and lazy updates

### 2020-02-20 paper/1.introduction.tex
fix based on the reviews

\noindent One feature of this composition is that $put_{bx_1 \circ bx_2}$ needs to call $get_{bx_1}$ to compute the intermediate result for $put_{bx_2}$ to use, which would introduce an efficiency problem if we compute $put$ for composition of many bidirectional programs. Generally, for a composition of $O(n)$ bidirectional programs, we need to call $get$ for $O(n^2)$ times. To be concrete, consider the evaluation of the following composition (which will be used as our running example in this paper):

@@

To treat large inputs, we propose the second method based on memoization: ``keeping \emph{complements in a closure} and using them when needed''. The key idea is straightforward: Complements are smaller
% program fragments
than
% the original
intermediate states.
%Therefore this strategy solves the previous two problems.
%Readers might already know that it is a hard problem to obtain complements in general. In our work, thanks to the following finding, it is not hard:
%
%\vspace{2mm}
%In very-well behaved \cite{Foster:2007:CBT:1232420.1232424} BX,% programs,
%$put$ is a complement function for $get$.
%, $get$ can be a complement function for $put$.
%\vspace{2mm}
%
For obtaining complements, we tuple $put$ and $get$, and produce a new function $pg$. Because $put$ produces new complements for $get$, we can shrink the size.
Let us reconsider our example $phead \circ phead \circ phead$ in Figure \ref{fig:eval-comp-phead-2}, where $c_1$ and $c_2$ are complements and $d_1$ and $d_2$ are valid views for $s$ and~$i_1$. Here, two points are worth noting. First, after evaluating the leftmost $pg$, the original source $s$ need not be kept because its complete contents are in $c_1$ and~$i_1$. Second, the complements are smaller than the intermediate states in Figure~\ref{fig:eval-comp-phead}.
%This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources.
Actually, this simple
% combined
$pg$ alone is not yet effective for left associative compositions because it requires two more $put$s, which can be seen on the right side of Figure \ref{fig:eval-comp-phead-2}. To achieve an efficient evaluation, we combine two techniques, lazy update and lazy evaluation. We explain this second method and all optimizations in Section~\ref{sect:xpg}.

>>>>

\noindent
\textcolor{red}{Note that the order of composition is the reverse order compared to function composition. We use this order, because it is standard in BX community.}
One feature of this composition is that $put_{bx_1 \circ bx_2}$ needs to call $get_{bx_1}$ to compute the intermediate result for $put_{bx_2}$ to use, which would introduce an efficiency problem if we compute $put$ for composition of many bidirectional programs. Generally, for a composition of $O(n)$ bidirectional programs, we need to call $get$ for $O(n^2)$ times. To be concrete, consider the evaluation of the following composition (which will be used as our running example in this paper):

@@

To treat large inputs, we propose the second method based on memoization: ``keeping \emph{complements in a closure} and using them when needed''.
\textcolor{red}{Here, complements are minimal parts that are needed to evaluate $put$. 
  In the middle $put$ and $get$ of Figure \ref{fig:eval-comp-phead-2},
  we use $i_1$, $[[1,2],[3]]$ and $i_2'$, $[100,2]$ to obtain the updated source $[[100,2],[3]]$. However, $[1,2]$ is simpley replaced by $[100,2]$ and not used to coustruct the result. In this case we can use $[..., [3]]$ as a complement.}
The key idea is straightforward: Complements are smaller
% program fragments
than
% the original
intermediate states.
%Therefore this strategy solves the previous two problems.
%Readers might already know that it is a hard problem to obtain complements in general. In our work, thanks to the following finding, it is not hard:
%
%\vspace{2mm}
%In very-well behaved \cite{Foster:2007:CBT:1232420.1232424} BX,% programs,
%$put$ is a complement function for $get$.
%, $get$ can be a complement function for $put$.
%\vspace{2mm}
%
For obtaining complements, we tuple $put$ and $get$, and produce a new function $pg$. Because $put$ produces new complements for $get$, we can shrink the size.
Let us reconsider our example $phead \circ phead \circ phead$ in Figure \ref{fig:eval-comp-phead-2}, where $c_1$ and $c_2$ are complements and $d_1$ and $d_2$ are valid views for $s$ and~$i_1$. Here, two points are worth noting. First, after evaluating the leftmost $pg$, the original source $s$ need not be kept because its complete contents are in $c_1$ and~$i_1$. Second, the complements are smaller than the intermediate states in Figure~\ref{fig:eval-comp-phead}.
%This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources.
Actually, this simple
% combined
$pg$ alone is not yet effective for left associative compositions because it requires two more $put$s, which can be seen on the right side of Figure \ref{fig:eval-comp-phead-2}. To achieve an efficient evaluation, we combine two techniques, lazy update and lazy evaluation. We explain this second method and all optimizations in Section~\ref{sect:xpg}.

### 2020-02-20 paper/4.xpg.tex
fix based on the reviews

A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling occurs, there will be 4 combinations from these conditions. This means two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satisfied. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v)) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper\footnote{\textcolor{red}{If the BX programs have efficient inverse way of evaluation, they are not affected by this restriction. Of course, some programs does not have such efficient way. An example restricted by this is evaluating $snoc$ with different length of a source and a view.}}. This does not happen for the others which is used in the construction of $\pginline{Case}$.

@@

\textcolor{red}{TODO-C6: RearrS, RearrV just keeping, Really evaluated is skip, product}
In $kpg$, basically, functions for the updates will be kept in $lks$ and $lkv$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $lks \ s$ and $lkv \ v$ and update $lks$ and $lkv$ with the identity functions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $\underline{lks_1}$ and $\underline{s_1}$ in the result of the first assignment are still redundant but not evaluated.
\\
Additionally we did two optimizations in $kpg$. The first is in $\kpginline{bx_1 \times bx_2}$. The evaluation of $lks$ and $lkv$ will be done independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be same computations in $fst \circ lks$ and $snd \circ lks$ as well as $fst \circ lkv$ and $snd \circ lkv$.
To remove duplicate evaluations, we evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$.
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$. We need to evaluate $lks \ s'$ and $lkv \ v'$ to check the \underline{fi} condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

>>>>

    A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling occurs, there will be 4 combinations from these conditions. This means two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satisfied. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v)) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper\footnote{\textcolor{red}{If the BX programs have efficient inverse way for evaluation, they are not affected by this restriction. Of course, all BX programs do not always have such efficient way. An example restricted by this is evaluating $bsnoc$ with different length of a source and a view.}}. This does not happen for the others which is used in the construction of $\pginline{Case}$.

@@

\textcolor{red}{
In $kpg$, basically, functions for the updates will be kept in $lks$ and $lkv$.
In $\kpginline{RearrS}$ and $\kpginline{RearrV}$, $f_1$ and $g_1$ are accumulated in $lks$ and $lkv$.
The kept functions are evaluated in $\kpginline{Skip}$ and $\kpginline{Case}$ by application of $lks \ s$ and $lkv \ v$. At the same time, $lks$ and $lkv$ are updated with the identity functions.
This evaluation is needed, because these definitions require the actual values, $es$ and $ev$.
Thanks to this accumulation in $kpg$, $\underline{lks_1}$ and $\underline{s_1}$ in $\kpginline{bx_1 \circ bx_2}$ are not evaluated as possible.}
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.}

\textcolor{red}{Additionally we did two optimizations in $kpg$. The first is in $\kpginline{bx_1 \times bx_2}$.
  Because $es$ and $ev$ are not used in this definition, we do not need to evaluate.
  However, if we accumulate $lks$ and $lkv$, both might be evlauated independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. This includes the same computation. To remove duplicate evaluations, we evaluate actual values $es$ and $ev$ before calling $\kpginline{bx_1}$. and $\kpginline{bx_2}$}
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$. We need to evaluate $lks \ s'$ and $lkv \ v'$ to check the \underline{fi} condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

### 2020-02-20 paper/5.experiments.tex
fix based on the reviews

We have fully implemented and tested all methods% 
%
\footnote{All experiments on macOS 10.14.6, processor Intel Core i7 (2.6~GHz), RAM 16~GB 2400~MHz DDR4, OCaml 4.07.1. The OCaml runtime system options and garbage collection parameters are set as default.}$^,$%HACK!
%
\footnote{Anonymous \url{https://mega.nz/\#F!ec8y2ITa!tLOHKkHgs_7fjSI8m-O26Q}}
%
described in the previous sections. Our target language is untyped. Some dummies used for $pg$, $cpg$, $kpg$ and $xpg$ are replaced with the current updated views. This helps a program in the $put$ direction valid.

@@

\textcolor{red}{TODO-D4: emphasize overhead of memoization. it is not for free.}
Figure \ref{fig:result} 
% illustrates 
shows the evaluation times for each of the seven test cases using the three methods: $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$.
We also did similar experiments with $pg$, $cpg$ and $kpg$, but their results are slower than the corresponding ones of $xpg$. The slowness was caused by the exponential number of $pg$ calls in the case of $pg$, redundant evaluations ($cpg$), and redundant overhead for constructing closures ($kpg$).
Therefore we simply omit these results in Figure \ref{fig:result}.
As we all know, $put$ works poorly for left associative compositions because of the number of reevaluated $get$s.
%The figure shows $put_m$ and $xpg$ are better than $put$ in some cases.
The left part of the figure contains tests using not-large inputs, and we see that $put_m$ is the fastest method for them. However if the input size is large enough, in the cases of the right part of the figure, $put_m$ will be slower quickly due to time for manipulating data in the table. At that time, $xpg$ is 
% proved to be
the most effective method.
\textcolor{red}{TODO-B2: add short evaluation result of memory.}

>>>>

We have fully implemented and tested all methods% 
%
\footnote{All experiments on macOS 10.14.6, processor Intel Core i7 (2.6~GHz), RAM 16~GB 2400~MHz DDR4, OCaml 4.07.1. The OCaml runtime system options and garbage collection parameters are set as default.}$^,$%HACK!
%
\footnote{\textcolor{red}{The implementation is available: in GitHub?} \url{https://mega.nz/\#F!ec8y2ITa!tLOHKkHgs_7fjSI8m-O26Q}}
%
described in the previous sections. Our target language is untyped. Some dummies used for $pg$, $cpg$, $kpg$ and $xpg$ are replaced with the current updated views. This helps a program in the $put$ direction valid.

@@

Figure \ref{fig:result} 
% illustrates 
shows the evaluation times for each of the seven test cases using the three methods: $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$.
We also did similar experiments with $pg$, $cpg$ and $kpg$, but their results are slower than the corresponding ones of $xpg$. The slowness was caused by the exponential number of $pg$ calls in the case of $pg$, redundant evaluations ($cpg$), and redundant overhead for constructing closures ($kpg$).
Therefore we simply omit these results in Figure \ref{fig:result}.
As we all know, $put$ works poorly for left associative compositions because of the number of reevaluated $get$s.
%The figure shows $put_m$ and $xpg$ are better than $put$ in some cases.
The left part of the figure contains tests using not-large inputs, and we see that $put_m$ is the fastest method for them. However if the input size is large enough, in the cases of the right part of the figure, $put_m$ will be slower quickly due to time for manipulating data in the table. At that time, $xpg$ is 
% proved to be
the most effective method.

\textcolor{red}{Note that this result focuses on the BX programs that includes many compositions. If the number of compositions are small in the BX programs, $put$ will be fastest, because of the overhead for memoization ($put_m$) and the overhead for keeping complements in closures ($xpg$).}

\textcolor{red}{TODO-B2: add short evaluation result of memory.}

### 2020-02-20 paper/6.related-work.tex
fix based on the reviews

Since the pioneering work of lens \cite{Foster:2007:CBT:1232420.1232424}, many BX languages have been proposed \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487,Buchmann:2018:BFI:3362232.3362263,Ko:2016:BFV:2847538.2847544,LeblebiciAS14,Samimi-Dehkordi18,Cicchetti2011,Hinkel:2019:CPB:3318595.3318634}.
%(memo: BXtend, eMoflon, EVL+Strace, JTL, NMF. SDMLib should be added?)
%based on several application scenarios and BX lenses works \cite{}.
Although much progress has been made on the semantics and correctness of BX programs for the past years, as far as we are aware, little work has been done on optimization of BX programs \cite{Horn:2018:IRL:3243631.3236769}. Anjorin et. al. introduces the first benchmark for BX languages and compared them \cite{Anjorin2019}, but a systematic improvement for practical implementation of BX languages is still missing. This paper shows the first attempt of improving efficiency of BX composition evaluation.

>>>>

Since the pioneering work of lens \cite{Foster:2007:CBT:1232420.1232424}, many BX languages have been proposed \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487,Buchmann:2018:BFI:3362232.3362263,Ko:2016:BFV:2847538.2847544,LeblebiciAS14,Samimi-Dehkordi18,Cicchetti2011,Hinkel:2019:CPB:3318595.3318634}.
%(memo: BXtend, eMoflon, EVL+Strace, JTL, NMF. SDMLib should be added?)
%based on several application scenarios and BX lenses works \cite{}.
Although much progress has been made on the semantics and correctness of BX programs for the past years, as far as we are aware, little work has been done on optimization of BX programs \cite{Horn:2018:IRL:3243631.3236769}. Anjorin et. al. introduces the first benchmark\footnote{\textcolor{red}{The BX programs in their benchmark are almost single BX programs. Because we focus on the BX programs that includes many compositions, their benchmark is not applicable for our evaluation.}} for BX languages and compared them \cite{Anjorin2019}, but a systematic improvement for practical implementation of BX languages is still missing. This paper shows the first attempt of improving efficiency of BX composition evaluation.

### 2020-02-20 paper/7.conclusion.tex
add todo list
fix based on the reviews

We will continue our work on the following three points.
First is overcoming our limitation in $xpg$. In $xpg$, there are two limitations: only for very-well-behaved programs and restriction for $\pginline{Case}$ in Section~\ref{sect:pg}. To be a practical evaluation method, an extension of the target language is needed. After extension of the language, we can evaluate more various programs for experimental results.
Second is introducing an automatic analysis about BX programs and inputs to choose the best evaluation method.
Currently, programmers have to choose the evaluation method by themselves based on their BX programs and inputs. If this analysis is achieved, we can reduce programmers' burden.
Third is introducing a type system to our target language, especially the datatype of sources and view. If we introduce this, we can avoid runtime errors like statically-typed functional languages. For this, we need to investigate more about $construct\_dummy$, because the current definition used in our experiments %Section~\ref{sect:experiments}
will cause type errors.

>>>>

We will continue our work on the following four points.
First is overcoming our limitation in $xpg$. In $xpg$, there are two limitations: only for very-well-behaved programs and restriction for $\pginline{Case}$ in Section~\ref{sect:pg}. To be a practical evaluation method, an extension of the target language is needed. After extension of the language, we can evaluate more various programs for experimental results.
Second is introducing an automatic analysis about BX programs and inputs to choose the best evaluation method.
Currently, programmers have to choose the evaluation method by themselves based on their BX programs and inputs. If this analysis is achieved, we can reduce programmers' burden.
Third is introducing a type system to our target language, especially the datatype of sources and view. If we introduce this, we can avoid runtime errors like statically-typed functional languages. For this, we need to investigate more about \textcolor{red}{how to construct dummy values}, because the current definition used in our experiments %Section~\ref{sect:experiments}
will cause type errors.
\textcolor{red}{Fourth is using a lazy language. Although we used a strict language OCaml in this paper, if we use a lazy language, we might skip one of the optimizations.}
